VOI-based Monte Carlo Sampling

### 2011-09-13 paper/rmcts/paper.tex
small fix
more papers

\begin{abstract}
UCT, a state-of-the art algorithm for Monte Carlo tree sampling
(MCTS), is based on UCB, a sampling policy for the Multi-armed Bandit
Problem (MAB) that minimizes the accumulated regret. However, MCTS
differs from MAB in that only the final choice, rather than all arm
pulls, brings a reward, that is, the simple regret, as opposite to the
cumulative regret, must be minimized. This work introduces policies for
multi-armed bandits with lower simple regret than UCB, and an
algorithm for MCTS which combines cumulative and simple regret
minization and outperforms UCT. Finite-time and asymptotic analysis of
the policies is provided, and the algorithms are empirically compared.
\end{abstract}

@@

The Multi-armed Bandit is a set of $K$ arms. Each arm can be pulled
multiple times. When the $i$th arm is pulled, a random reward $X_i$
from an unknown stationary distribution is returned.  The reward is 
bounded between 0 and 1.

The simple regret of a sampling policy for the Multi-armed Bandit
Problem is the expected difference between the best expected reward
$\mu_*$ and the expected reward $\mu_j$ of the arm with the best sample mean
$\overline X_j=\max_i\overline X_i$:
\begin{equation}
\label{eq:simple-regret}
\IE[R]=\sum_{j=1}^K\Delta_j\Pr(\overline X_j=\max_i\overline X_i)
\end{equation}
where $\Delta_j=\mu_*-\mu_j$.
Strategies that minimize the simple reward are called pure exploration
strategies \cite{Bubeck.pure}.

Such strategies are used to select the best arm, and, by extension,
the best action in Monte Carlo tree search. Monte Carlo tree search is
used to solve Markov Decision Processes (MDP) approximately. An MDP is
defined by the set of states $S$, the set of actions $A$, the
transition table $T(s, a, s')$, the reward table $R(s, a, s')$, the
initial state $s$ and the goal state $t$: $(S, A, T, R, s, t)$
\cite{Russell.aima}.  Monte Carlo tree search explores an MDP by
\emph{rollouts}---trajectories from the current state to a state in
which a termination condition is satisfied (either the goal state, or
a cutoff state for which the reward is evaluated
approximately). Algorithm UCB that minimizes the cumulative regret
\cite{Auer.ucb} had been extended to tree search as UCT
\cite{Kocsis.uct}.

UCT-driven search achieves good results in many search problems. At each
search step, the algorithm allocates Monte Carlo rollouts to choose
the best action, thereby minimizing the cumulative regret. However, the
simple and the cumulative regret cannot be minimized simultaneously;
moreover, \cite{Bubeck.pure} shows that in many cases the smaller the
cumulative regret, the greater the simple regret.

UCT performance can be improved by combining UCB with a sampling
scheme that minimizes the simple regret of selecting an action at the
current root node. Indeed, the algorithm must select an action with
the minimum regret \emph{given the assumption that after performing the selected action
the algorithm performs optimally}, which corresponds to maximizing the
value of partial information \cite{Russell.aima}. Therefore, an
improved allocation scheme would
\begin{itemize}
\item maximize the value of partial
information by sampling actions to minimize the \textbf{simple} regret at the
current root node, and then 
\item minimize the \textbf{cumulative} regret of the rollouts from the second
  step on to approach the optimum behavior. 
\end{itemize}

@@

\begin{enumerate}
\item The $\varepsilon$-greedy allocation scheme has exponential convergence
rate of the simple regret $\IE r$ (see Appendix~\ref{app:derivations-finite-eps}):
\begin{equation}
  \IE r_\varepsilon\le\sum_{j=1}^K\Delta_j\left(\frac {n\varepsilon} K + \frac {4\sqrt e}
{\Delta_j^2}\right)e^{-\Delta_j^2n\varepsilon/8K}
\end{equation}
\item The UCB allocation scheme exhibits polynomial convergence
rate (see Appendix~\ref{app:derivations-finite-ucb}):
\begin{equation}
\IE r_{ucb} \le 2\sum_{j=1}^K \Delta_jn^{\frac {-\alpha \Delta_j^2} 8}
\end{equation}
\item A tighter upper regret bound for the $\varepsilon$-greedy allocation scheme depends on
$\varepsilon$. The lowest bound is
acheived for $\varepsilon=\frac 1
2$, and for a large number of arms the bound for the $\frac 1 2$-greedy scheme approaches the square of the
bound for uniform sampling (see
Appendix~\ref{app:derivations-asym-eps}).

@@

Figure~\ref{fig:sailing-cost-vs-factor} shows the rerget vs. the
exploration factor for different numbers of samples. QCT is always better than
UCT, and $\frac 1 2$-greedy is better than UCT expect for a small range of
values of the exploration factor $C_p$.

@@

The improvement is inspired by the notion of value of information (VOI),
but VOI is used implicitly in the analysis of the algorithm, rather
than computed or learned explicitly in order to plan the rollouts. A
further improvement can be achieved by computing or estimating the VOI
of the rollouts and chosing a rollout that maximizes the VOI. Using VOI
to guide Monte Carlo sampling instead of relying on the sample means
of different actions  is particularly beneficial when different
actions have qualitatively different distributions: for example, when the
distribution variance varies signficantly between the actions.

@@

VOI of a rollout can be computed when the sample distribution of an
action is known up to the parameters, such as the normal distribution
with an unknown mean and/or variance. Alternatively, the value of
information can be estimated from the set of samples, and the need to
assume a particular shape of the distribution is be lifted. In one
realization of the latter approach the VOI of performing an action is
estimated as \emph{the value of perfect information learned from the outcome
of earlier samples of the action divided by the number of
samples}. Early experiments with this approach showed the best
performance for sets of Bernoulli arms, and even a more prominent
improvement for mixed arms (e.g. Bernoulli, triangularly distributed,
and fixed arms) (Figure~\ref{fig:flat-trilevel-vct}):
\textbf{VCT}, the algorithm which selects an action with the maximum VOI estimate on
the first step exhibits lower average regret than other algorithms.
Sample-based VOI estimation requires keeping more information than an
algorithm based only on the sample
mean, but the overhead is small and
is justified by the improvement.

@@

Bounds (\ref{eq:deriv-epsilon-greedy-b}, \ref{eq:deriv-ucb-alpha-b}) were
derived ignoring dependency of the allocation on expected values
of non-optimal arms. Derivation of better bounds seems
complicated. Instead, one can analyse the asymptotic behavior of each of the
algorithms under the assumption that the sample mean is close to the expectation.

@@

Splitting the interval between $\mu_j$ and $\mu_*$ half-way is an
arbirtrary decision. By selecting a different split point
depending on the allocation of samples, one can obtain a tighter
bound. On the other hand, there is a value of $\varepsilon$ that
minimizes the bound for the $\varepsilon$-greedy allocation scheme.
Thus, the  $\varepsilon$-greedy allocation scheme can be tuned by
finding $\varepsilon$ that minimizes the bound obtained by
selecting the best split point.

>>>>

\begin{abstract}
UCT, a state-of-the art algorithm for Monte Carlo tree sampling
(MCTS), is based on UCB, a sampling policy for the Multi-armed Bandit
Problem (MAB) that minimizes the accumulated regret. However, MCTS
differs from MAB in that only the final choice, rather than all arm
pulls, brings a reward, that is, the simple regret, as opposite to the
cumulative regret, must be minimized. This ongoing work aims at applying
meta-reasoning techniques to MCTS, which is non-trivial.
We begin by introducing policies for
multi-armed bandits with lower simple regret than UCB, and an
algorithm for MCTS which combines cumulative and simple regret
minimization and outperforms UCT. We also develop a sampling scheme loosely based
on a myopic version of perfect value of information.
Finite-time and asymptotic analysis of
the policies is provided, and the algorithms are compared empirically.
\end{abstract}

@@

Monte-Carlo tree sampling, and especially a version based on the
UCT formula \cite{Kocsis.uct} appears in numerous search applications,
such as \cite{Eyerich.ctp}. Although these methods are shown to be successful empirically,
most authors appear to be using the UCT formula ``because it has been shown
to be successful in the past'', and ``because it does a good job of
trading off exploration and exploitation''. While the latter statement may be
correct for the multi-armed bandit and for the UCB method \cite{Auer.ucb},
we argue that it is inappropriate for search. The problem is not that
UCT does not work; rather, we argue that a simple reconsideration from basic
principles can result in schemes that outperform UCT. This is demonstrated both
theoretically and empirically.

In the Multi-armed Bandit problem we have a set of $K$ arms. Each arm can be pulled
multiple times. When the $i$th arm is pulled, a random reward $X_i$
from an unknown stationary distribution is returned.  The reward is 
bounded between 0 and 1.

The simple regret of a sampling policy for the Multi-armed Bandit
Problem is the expected difference between the best expected reward
$\mu_*$ and the expected reward $\mu_j$ of the arm with the best sample mean
$\overline X_j=\max_i\overline X_i$:
\begin{equation}
\label{eq:simple-regret}
\IE[R]=\sum_{j=1}^K\Delta_j\Pr(\overline X_j=\max_i\overline X_i)
\end{equation}
where $\Delta_j=\mu_*-\mu_j$.
Strategies that minimize the simple regret are called pure exploration
strategies \cite{Bubeck.pure}.

Such strategies are used to select the best arm, and, by extension,
the best action in Monte Carlo tree search. Monte Carlo tree search is
used to solve Markov Decision Processes (MDP) approximately. An MDP is
defined by the set of states $S$, the set of actions $A$, the
transition table $T(s, a, s')$, the reward table $R(s, a, s')$, the
initial state $s$ and the goal state $t$: $(S, A, T, R, s, t)$
\cite{Russell.aima}.  Monte Carlo tree search explores an MDP by
\emph{rollouts}---trajectories from the current state to a state in
which a termination condition is satisfied (either the goal state, or
a cutoff state for which the reward is evaluated
approximately). The UCB algorithm (that attempts to minimize the cumulative regret)
\cite{Auer.ucb} had been extended into the tree search sampling scheme
known as UCT \cite{Kocsis.uct}.

UCT-driven search achieves good results in many search problems. At each
search step, the algorithm allocates Monte Carlo rollouts to choose
the best action, thereby (approximately) minimizing the cumulative regret. 
The core issue is that in search (especially for adversarial search
and for ``games against nature'' - optimizing behavior under uncertainty) the goal is typically
to either find a good (or optimal) strategy, or even to find the best first
action of such a policy. Once such an action is discovered, it is not beneficial
to further sample that action, ``exploitation'' is thus meaningless for
search problems. Finding a good first action is closer to the pure exploration
variant, as seen in the selection problem \cite{Bubeck.pure,TolpinShimony.blinkered}. In the selection
problem, it is much better to minimize the {\em simple} regret.
However, the simple and the cumulative regret cannot be minimized simultaneously;
moreover, \cite{Bubeck.pure} shows that in many cases the smaller the
cumulative regret, the greater the simple regret.

UCT performance can thus be improved by combining UCB with a sampling
scheme that minimizes the simple regret of selecting an action at the
current root node. Indeed, the algorithm must select an action with
the minimum regret \emph{given the assumption that after performing the selected action
the algorithm performs optimally}, which corresponds to maximizing the
value of partial information \cite{Russell.aima}. Therefore, an
improved allocation scheme would
\begin{itemize}
\item maximize the value of partial
information by sampling actions to
minimize the \textbf{simple} regret
of the selection at the current root node, and
\item as the goal of sampling in deeper tree nodes is estimating the value of
a node, rather than selection, it makes sense to
minimize the \textbf{cumulative} regret of the rollouts from the second
step onwards. 
\end{itemize}

The ultimate goal of this ongoing work is ``optimal'' sampling
in the meta-reasoning sense. Nevertheless, this task is daunting for
the following reasons:
\begin{itemize}
\item Defining the cost of a sample is not easy, and even if we simply use time-cost
as an approximation, we get an intractable meta-reasoning problem.
\item Applying the standard myopic and subtree independence assumptions, we run into
serious problems. Even in the standard selection problem \cite{TolpinShimony.blinkered}, we get a non-concave
utility function and premature stopping of the algorithm. This is due to the fact
that the value of information of a single measurement (analogous
to a sample in MCTS) is frequently less than its
time-cost, even though this is not true for multiple measurements.
When applying the selection problem to MCTS, the situation is exacerbated. 
The utility of an action is usually bounded, and thus in many cases a single sample
may be insufficient to change the current best action, {\em regardless} of its
outcome. As a result, we frequently
get a {\em zero} ``myopic'' value of information for a single sample.
\end{itemize}

As the ultimate task is extremely difficult
to achieve, and even harder to analyze, we begin with
simple schemes more amenable to analysis. We then
develop a crude value of information (VOI)
based scheme. Although the latter is a somewhat ad-hoc attempt to
estimate value of information of a sample, it already appears (empirically) to do
better than UCT as well as better than all the simpler schemes we propose.

@@

\begin{enumerate}
\item The $\varepsilon$-greedy allocation scheme has exponential convergence
rate of the simple regret $\IE r$ (see Appendix~\ref{app:derivations-finite-eps}):
\begin{equation}
  \IE r_\varepsilon\le\sum_{j=1}^K\Delta_j\left(\frac {n\varepsilon} K + \frac {4\sqrt e}
{\Delta_j^2}\right)e^{-\Delta_j^2n\varepsilon/8K}
\end{equation}
\item The UCB allocation scheme exhibits polynomial convergence
rate (see Appendix~\ref{app:derivations-finite-ucb}):
\begin{equation}
\IE r_{ucb} \le 2\sum_{j=1}^K \Delta_jn^{\frac {-\alpha \Delta_j^2} 8}
\end{equation}
\item A tighter upper regret bound for the $\varepsilon$-greedy allocation scheme depends on
$\varepsilon$. The lowest bound is
achieved for $\varepsilon=\frac 1
2$, and for a large number of arms the bound for the $\frac 1 2$-greedy scheme approaches the square of the
bound for uniform sampling (see
Appendix~\ref{app:derivations-asym-eps}).

@@

Figure~\ref{fig:sailing-cost-vs-factor} shows the regret vs. the
exploration factor for different numbers of samples. QCT is always better than
UCT, and $\frac 1 2$-greedy is better than UCT expect for a small range of
values of the exploration factor $C_p$.

@@

The improvement is inspired by the notion of value of information (VOI),
but VOI is used implicitly in the analysis of the algorithm, rather
than computed or learned explicitly in order to plan the rollouts. A
further improvement can be achieved by computing or estimating the VOI
of the rollouts and choosing a rollout that maximizes the VOI. Using VOI
to guide Monte Carlo sampling instead of relying on the sample means
of different actions  is particularly beneficial when different
actions have qualitatively different distributions: for example, when the
distribution variance varies significantly between the actions.

@@

VOI of a rollout can be computed when the sample distribution of an
action is known up to the parameters, such as the normal distribution
with an unknown mean and/or variance. Alternatively, the value of
information can be estimated from the set of samples, and the need to
assume a particular shape of the distribution can be lifted. In one
realization of the latter approach the VOI of performing an action is
estimated as \emph{the value of perfect information learned from the outcome
of earlier samples of the action divided by the number of
samples}. Early experiments with this approach showed the best
performance for sets of Bernoulli arms, and even a more prominent
improvement for mixed arms (e.g. Bernoulli, triangularly distributed,
and fixed arms) (Figure~\ref{fig:flat-trilevel-vct}):
\textbf{VCT}, the algorithm which selects an action with the maximum VOI estimate on
the first step exhibits lower average regret than other algorithms.
Sample-based VOI estimation requires keeping more information than an
algorithm based only on the sample
mean, but the overhead is small and justified by the improvement.

@@

Bounds (\ref{eq:deriv-epsilon-greedy-b}, \ref{eq:deriv-ucb-alpha-b}) were
derived ignoring dependency of the allocation on expected values
of non-optimal arms. Derivation of better bounds seems
complicated. Instead, one can analyze the asymptotic behavior of each of the
algorithms under the assumption that the sample mean is close to the expectation.

@@

Splitting the interval between $\mu_j$ and $\mu_*$ half-way is an
arbitrary decision. By selecting a different split point
depending on the allocation of samples, one can obtain a tighter
bound. On the other hand, there is a value of $\varepsilon$ that
minimizes the bound for the $\varepsilon$-greedy allocation scheme.
Thus, the  $\varepsilon$-greedy allocation scheme can be tuned by
finding $\varepsilon$ that minimizes the bound obtained by
selecting the best split point.

### 2011-09-13 paper/voisam/paper.tex
more papers

\section{Introduction}

>>>>

\section{Introduction and Definitions}

Taking a sequence of samples in order to minimize the
regret of a decision based on the samples is abstracted by the
{\em Multi-armed Bandit Problem.} In the Multi-armed Bandit problem
we have a set of $K$ arms. Each arm can be pulled multiple
times. When the $i$th arm is pulled, a random reward $X_i$ from an
unknown stationary distribution is returned.  The reward is bounded
between 0 and 1.

The simple regret of a sampling policy for the Multi-armed Bandit
Problem is the expected difference between the best expected reward
$\mu_*$ and the expected reward $\mu_j$ of the arm with the best sample mean
$\overline X_j=\max_i\overline X_i$:
\begin{equation}
\label{eq:simple-regret}
\IE[R]=\sum_{j=1}^K\Delta_j\Pr(\overline X_j=\max_i\overline X_i)
\end{equation}
where $\Delta_j=\mu_*-\mu_j$.
Strategies that minimize the simple regret are called pure exploration
strategies \cite{Bubeck.pure}. Principles of rational metareasoning
\cite{Russel.right} suggest that at each step the arm with the great
value of information (VOI) must be pulled, and the sampling must be
stopped and a decision must be made when no arm has positive VOI.

To estimate the VOI of pulling an arm, either a particular 
distribution of the rewards should be assumed (and updated based on
pull outcomes), or a distribution-independent bound on the VOI can be
used as the estimate. In this paper, we use {\em concentration inequalities}
to derive distribution-independent bounds on the VOI.

\section{Some Concentration Inequalities}

\begin{description}
\item[Markov' inequality \cite{}:] Let $X \ge 0$ be a random non-negative variable
  with finite mean $\IE[X]$. Then for all $a > 0$
\begin{equation}
\Pr(X\ge a)\le \frac {\IE[X]} a
\label{eq:conc-markov-upper}
\end{equation}
If additionally $X\le 1$, then by susbstituting $Y=1-X$
\begin{equation}
\Pr(X\le a)\le \frac {1-\IE[X]} {1-a}
\label{eq:conc-markov-lower}
\end{equation}
\item[Hoeffding's inequality \cite{}:] Let $X_1, \ldots, X_n$ be
independent random variables with values from $[0,1]$,
$X=\frac 1 n \sum_{i=1}^n X_i$. Then 
\begin{eqnarray}
\Pr(X-\IE[X] \ge a) &\le& \exp ( -2na^2)\nonumber\\
\Pr(\IE[X]-X \ge a) &\le& \exp ( -2na^2)
\label{eq:conc-hoeffding}
\end{eqnarray}
\item[Maurer \& Pontil's inequality \cite{MaurerPontil.benrstein}:] \footnote{A
  modification of Bennet's inequality \cite{} based on the sample,
  rather than the true, variance.}
\end{description}

\section{Upper Bounds on Value of Information}

\section{VOI-based Sampling Control}

\subsection{Selection Criterion}

\subsection{Termination Condition}
